---
title: "ga_optimizer"
author: "Jasleen Grewal"
date: '2016-10-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GA algorithm
This is an R package that uses genetic algorithms for function parameter optimzation <https://cran.r-project.org/web/packages/GA/index.html>.

```{r loadlibs, echo=FALSE}
library("testthat")
library("GA")
library("readr")
```

##Source data reading functions 
Let us source the wrapper functions for our tool (Abyss)

```{r getwrapper, echo=FALSE,eval=FALSE}
#TO DO: have a function here that wgets the data file and points infile var to downloaded file
#setwd("/Users/jgrewal/mygits/2016_project_2/")
setwd("/Users/jgrewal/mygits/common/hackseq_2016/")
source(paste(getwd(), "scripts/R_optimizers/common/AbyssWrapper.R", sep="/"))
```

##Define function to optimize
```{r funcopt, echo=TRUE}
#Target metric is N50
#f <- function(x)  (return(runAbyssTest(k=round(x))))
f <- function(x) (x-30)^2
f_twoparam <- function(x1, x2) (x1 + x2)
```

##Define test parameter ranges
```{r params, echo=TRUE}
getrange <- function(min, max, stepsize=0){#If stepsize zero, range is continuous
  real_val='real-valued'
  if(!(stepsize)){stepsize=max-min; real_val=FALSE}
  return(list(seqdat=seq(min, max, stepsize), seqtype=real_val))
  }
k = list(min=16, max=49, seqtype='real-valued')
np = list(min=10, max=20, seqtype='permutation')
multiparam = list(k, np)
```

##1. Call GA algorithm (single param fit)
```{r gafunc, echo=FALSE,results='hide'}
#type = c("binary", "real-valued", "permutation")
##type.real-valued = decision variables are floating-point representations of real numbers. 
##type.permutation = for problems that involves reordering of a list.
#optim = False (default), doing local optimization search
#optimArgs = c(method="L-BFGS-B") (default) 
inparam=k
GA <- ga(type=inparam$seqtype, fitness=f, min=inparam$min, max=inparam$max, monitor=TRUE)
```
View summary of fit
```{r gafit}
summary(GA)
```
Plot path of search
```{r searchpath, echo=FALSE}
plot.ga-method(GA)

```
Now let us visualize the GA results across a single parameter fit
```{r plotga, echo=FALSE}
plot(GA)
```

###Response variable vs Input parameter plot
```{r varplot, echo=FALSE}
curve(f, k$min, k$max, n=1000)
points(GA@solution, GA@fitnessValue,col=2,pch=19)
```

##2. Multi-metric optimization  
(other metric is L50)  

##3. Call GA algorithm (two param fit)
```{r gafunctwoparam, results='hide'}
#type can only be one type 
##unlist(lapply(multiparam,function(x) (x$seqtype))) does not work
x1 = seq(multiparam[[1]]$min, multiparam[[1]]$max,1)
x2 = seq(multiparam[[2]]$min, multiparam[[2]]$max,1)

twofit = outer(x1, x2, f_twoparam)
GA_multi <- ga(type='real-valued', fitness=function(x)(f_twoparam(x[1],x[2])), min=unlist(lapply(multiparam,function(x) (x$min))), max=unlist(lapply(multiparam,function(x) (x$max))), monitor=TRUE)
```
Display summary of fit
```{r gafunctwoparamsum}
summary(GA_multi)
plot(GA)
```

###Visualize param1 vs param2 with optimal combination indicated
```{r 3dplotresult}
#persp3D(multiparam[1]$seqtype, multi)
filled.contour(x1,x2, twofit, color.palette = bl2gr.colors, 
  plot.axes = { axis(1); axis(2); 
                points(GA_multi@solution[,1], GA_multi@solution[,2], 
                       pch = 3, cex = 2, col = "white", lwd = 2) }
)
```

###Pareto Frontier plot 
Response variable 1 vs Response variable 2
```{r makepareto}

```
