---
title: "genalg_optimizer"
author: "Jasleen Grewal"
date: '2016-10-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Package genalg contains rbga(), an implementation of a genetic algorithm for multi-dimensional function optimization.
```

## GA algorithm
This is an R package that uses genetic algorithms for function parameter optimzation <https://cran.r-project.org/web/packages/GA/index.html>.

```{r loadlibs, echo=FALSE}
library("testthat")
library("genalg")
library("readr")
```

##Source data reading functions 
Let us source the wrapper functions for our tool (Abyss)

```{r pressure, echo=FALSE}
#TO DO: have a function here that wgets the data file and points infile var to downloaded file
setwd("/Users/jgrewal/mygits/2016_project_2/")
source(paste(getwd(), "scripts/R_optimizers/common/AbyssWrapper.R", sep="/"))
```

##Define function to optimize
```{r funcopt, echo=TRUE}
#Target metric is N50
#f <- function(x)  (return(runAbyssTest(k=round(x))))
f <- function(x) (x-30)^2

```

##Define test parameter ranges
```{r params, echo=TRUE}
getrange <- function(min, max, stepsize=0){#If stepsize zero, range is continuous
  real_val='real-valued'
  if(!(stepsize)){stepsize=max-min; real_val=FALSE}
  return(list(seqdat=seq(min, max, stepsize), seqtype=real_val))
  }
k = list(min=16, max=49, seqtype='real-valued')
np = list(min=10, max=20, seqtype='permutation')
multiparam = list(k, np)
```

##Set up monitoring functions (single and multiple params)
```{r monitor_genalg, echo=FALSE}
devtools::install_github("dgrtwo/gganimate"); library(gapminder)
library(ggplot2)
theme_set(theme_bw())
```
##1Call genalg algorithm (single param fit)
```{r genalg_func, echo=FALSE,results='hide'}
#optim = False (default), doing local optimization search
#THIS PACKAGE IS A MINIMIZATION OPTIMIZER!!!
inparam=np


monitor_1param <- function(obj) {
  # plot the population
  itercount <<- itercount+1
  if(itercount==1){
    monitordat=data.frame(); 
  }
  monitordat <<- rbind(monitordat, data.frame(xmin=obj$stringMin[1],xmax=obj$stringMax[1], input_param=obj$population[,1],response_metric=obj$evaluations, iter=itercount))
  print(dim(monitordat));# print(obj);stop()
  xlim = c(obj$stringMin[1], obj$stringMax[1]);
  ylim = c(obj$stringMin[2], obj$stringMax[2]);
  plot(obj$population, xlim=xlim, ylim=ylim,
       xlab="input_param", ylab="response_metric"); 
}
itercount=0;
rbga.results = rbga(c(0, 1), c(inparam$max, 10), monitorFunc=monitor_1param,popSize=100,
                    evalFunc=function(x) abs(round(x[1])-5) + abs(round(x[2])-4),iter=100, verbose=TRUE, mutationChance=0.1)
plot(rbga.results); 

rbga.results = rbga(c(0),c(10), popSize=100,evalFunc=function(x) return(c(abs(round(x[1])-5),abs(round(x[1])-2))),iter=100, verbose=TRUE, mutationChance=0.1)

p <- ggplot(monitordat, aes(input_param, response_metric, frame = iter)) +geom_point() +scale_x_log10()
```

View summary of fit
```{r gafit}
cat(summary(rbga.results))
resultdat <- capture.output(cat(summary(rbga.results)))
bestsol= strsplit(resultdat[grep("Best Solution",resultdat)], split=" ")[[1]]
bestsol= as.numeric(bestsol[grep("[0-9]+", bestsol)])
bestsol_discrete = unlist(lapply(bestsol,round,0))
cat("Best input params are ", bestsol_discrete)

library(gganimate)
gg_animate(p, "output.gif")
plot(rbga.results)
```

###Response variable vs Input parameter plot
```{r varplot, echo=FALSE}
curve(f, k$min, k$max, n=1000)
points(GA@solution, GA@fitnessValue,col=2,pch=19)
```

##Call GA algorithm (two param fit)
```{r gafunctwoparam, results='hide'}
#type can only be one type 
##unlist(lapply(multiparam,function(x) (x$seqtype))) does not work
x1 = seq(multiparam[[1]]$min, multiparam[[1]]$max,1)
x2 = seq(multiparam[[2]]$min, multiparam[[2]]$max,1)

twofit = outer(x1, x2, f_twoparam)
GA_multi <- ga(type='real-valued', fitness=function(x)(f_twoparam(x[1],x[2])), min=unlist(lapply(multiparam,function(x) (x$min))), max=unlist(lapply(multiparam,function(x) (x$max))), monitor=TRUE)
```
Display summary of fit
```{r gafunctwoparamsum}
summary(GA_multi)
plot(GA)
```

###Visualize param1 vs param2 with optimal combination indicated
```{r 3dplotresult}
#persp3D(multiparam[1]$seqtype, multi)
filled.contour(x1,x2, twofit, color.palette = bl2gr.colors, 
  plot.axes = { axis(1); axis(2); 
                points(GA_multi@solution[,1], GA_multi@solution[,2], 
                       pch = 3, cex = 2, col = "white", lwd = 2) }
)
```

###Pareto Frontier plot 
Response variable 1 vs Response variable 2
```{r makepareto}

```
